Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-02-09 ThesisJP/Background/background.tex
Added new citation. Some small adjustments.

Around the same time as CPU manufactures hit the frequency wall people started experimented with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had previously been a fixed function pipeline, where e.g. vertices and texture coordinates where fed to the graphics processing unit (GPU) in one end and a rasterized rendered image was outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics. The graphics pipeline could for instance be hacked to evaluate a differential equation over grid defined by graphics primitives\todo{cite Seland et al.}. The rationale behind this exploration was that hidden in the GPU was processing power an order of magnitude larger than what the CPU could provide at the same time. Actually, at this time, the GPU had already gone multi-core, driven by the increasing demand for realistic computer games. Multiple cores was somewhat easier to achieve with the GPU since the rendering problem is inherently parallel. The pixels in the rendered image can be calculated in any order, and one pixel is therefore an concealed unit which the GPU scheduler is free to issue at any time to whatever core that is available. It it is also much less costly to switch between a pixel-thread for the GPU then a program thread for the CPU. Since there was only one program running on the GPU, the fixed function pipeline, GPU designers had a good idea of how the program would execute, and could therefore skip advance feature as predicting execution paths and different levels of data caching.  The released silicon was used to make more cores.

@@

\begin{figure}
\centering
\subfigure[Threads, blocks and grid]{
	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}
}
\subfigure[Kernel]{
	%\includegraphics[width=0.5\textwidth]{img/kernel.pdf}
}
\caption{Depicts how GPU threads are grouped into blocks and arranged in a grid. One thread runs a copy of a kernel function, in this example an element-wise matrix square operation.}
\label{fig:gpu_grid}
\end{figure}

>>>>

Around the same time as CPU manufactures hit the frequency wall people started experimented with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had previously been a fixed function pipeline, where e.g. vertices and texture coordinates where fed to the graphics processing unit (GPU) in one end and a rasterized rendered image was outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this exploration was that hidden in the GPU was processing power an order of magnitude larger than what the CPU could provide at the same time. Actually, at this time, the GPU had already gone multi-core, driven by the increasing demand for realistic computer games. Multiple cores was somewhat easier to achieve with the GPU since the rendering problem is inherently parallel. The pixels in the rendered image can be calculated in any order, and one pixel is therefore an concealed unit which the GPU scheduler is free to issue at any time to whatever core that is available. It it is also much less costly to switch between a pixel-thread for the GPU then a program thread for the CPU. Since there was only one program running on the GPU, the fixed function pipeline, GPU designers had a good idea of how the program would execute, and could therefore skip advance feature as predicting execution paths and different levels of data caching.  The released silicon was used to make more cores.

@@

\begin{figure}
\centering
\subfigure[Threads, blocks and grid]{
	\includegraphics[width=0.4\textwidth]{img/cuda_threads.png}
}
\subfigure[Kernel]{
	\includegraphics[width=0.5\textwidth]{img/kernel.pdf}
}
\caption{Depicts how GPU threads are grouped into blocks and arranged in a grid. One thread runs a copy of a kernel function, in this example an element-wise matrix square operation.}
\label{fig:gpu_grid}
\end{figure}

### 2014-02-09 ThesisJP/Thesis.tex
Added symbol list. Started writing the introduction.
Adjusted title.
Writing. Still some sections unfinished in discussion part.
Started adding figures in background section.
STarted writing about gpgpu.
Started adding references.
Added new citation. Some small adjustments.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TITLE PAGE
\author{Jon Petter \AA{}sen}
\title{%Integrated Ultrasound Visualization and Formation
%Real-Time Adaptive Beamforming, Visualization, and Simulation 
%Real-Time Adaptive Ultrasound Algorithms
Accelerating adaptive ultrasound algorithms using graphics processing units 
\vspace{0.5cm}}
\department{Department of Circulation and Medical Imaging}

@@

\begin{document}
\date{\today}
\pagestyle{empty} \maketitle

@@

This thesis is based on the following five papers, referred to in the text by their Roman numerals (I-V): 
\begin{enumerate}[I]
\renewcommand\labelenumi{\bfseries\theenumi}
	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "An optimised GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, submitted}.
	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon beamforming on a GPU for real-time cardiac ultrasound imaging", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. \todo{add pages} Jan 2014}.
 	\item \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm, "Capon beamforming and moving objects - An analysis of lateral shift-invariance", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}.
	\item \textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben, "Adaptive volume rendering of cardiac 3D ultrasound images - utilizing blood pool statistics", {\it Proc. SPIE Medical Imaging 2012 \todo{add pages}}.
	\item \textbf{J.\:P.\:\AA{}sen} and S.\:Holm, "Huygens on speed: Interactive simulation of ultrasound pressure fields", {\it Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}}.
\end{enumerate}

@@

	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon Beamforming on the GPU for Real-Time Cardiac Ultrasound Imaging", {\it Proc. IEEE Ultrasonics Symposium 2012} \todo{add pages}.

	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "Adapting the Minimum variance beamformer to a graphics processing unit for active sonar imaging systems", {\it Proc. Meetings on Acoustics 2013} \todo{add pages?}.

	\item G.\:Kiss, E.\:Steen, \textbf{J.\:P.\:\AA{}sen} and H.\:Torp, "GPU volume rendering in 3D echocardiography: real-time pre-processing and ray-casting", {\it Proc. IEEE Ultrasonics Symposium 2010} \todo{add pages}.

@@

	\item E.\:S.\:Br\o{}nstad, \textbf{J.\:P.\:\AA{}sen}, H.\:Torp and G.\:Kiss, "Visibility driven visualization of 3D cardiac ultrasound data on the GPU", {\it Proc. IEEE Ultrasonics Symposium 2012} \todo{add pages}.
\end{enumerate}

@@

\clearpage
\setcounter{page}{1}
\renewcommand{\thepage}{\arabic{page}}

\renewcommand*\thesection{\arabic{section}}

%%%% CHAPTER 1 %%%%
\begin{bibunit}[ieeetr]
\CustChapter*{Introduction} 
\addcontentsline{toc}{chapter}{Introduction}
\runningtitle{Introduction}
\input{./Introduction/introduction}
\newpage\addcontentsline{toc}{section}{References}
\putbib[bibIntroduction]
\end{bibunit}

\setcounter{section}{0}
\setcounter{chapter}{1}

%%%% CHAPTER 2 %%%%
\begin{bibunit}[ieeetr]
\CustChapter*{Background}
\addcontentsline{toc}{chapter}{Background} 
\runningtitle{Background}
\input{./Background/background}
\newpage\addcontentsline{toc}{section}{References}
\putbib[bibBackground]
\end{bibunit}

@@

%%%% Paper 1 %%%%
\begin{bibunit}[ieeetr]
\CustChapterPaper{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
\runningtitle{GPU MVDR Beamformer for Active Sonar Imaging}
\authors{
	Jo Inge Buskenes$^{1}$, 
    	\textbf{Jon Petter \AA{}sen}$^{2}$, 
    	Carl-Inge C. Nilsen$^{1}$ and
    	Andreas Austeng$^{1}$
} 
{
	$^{1}$Department of Informatics, University of Oslo, Oslo, Norway\\
    	$^{2}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway
}
\noindent \textit{IEEE Transactions on Oceanic Engineering, submitted}
%\newpage\input{./CaponSonarGPU/CaponSonarGPU}
%\newpage\addcontentsline{toc}{section}{References}
\putbib[bibCaponSonarGPU]
\end{bibunit}

%%%% Paper 2 %%%%
\begin{bibunit}[ieeetr]
\CustChapterPaper{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
\runningtitle{GPU Capon Beamformer for Cardiac Ultrasound Imaging}
\authors{
	\textbf{Jon Petter \AA{}sen}$^{1}$,
    	Jo Inge Buskenes$^{2}$,  
    	Carl-Inge C. Nilsen$^{2}$,
    	Andreas Austeng$^{2}$ and
    	Sverre Holm$^{1,2}$
} 
{
	$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
    	$^{2}$Department of Informatics, University of Oslo, Oslo, Norway
}
\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, Jan 2014}
%\input{./CaponCardiacGPU/CaponCardiacGPU}
%\newpage\addcontentsline{toc}{section}{References}
\putbib[bibCaponCardiacGPU]
\end{bibunit}

%%%% Paper 3 %%%%
\begin{bibunit}[ieeetr]
\CustChapterPaper{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}
\runningtitle{Capon Beamforming and Moving Objects}
\authors{
	\textbf{Jon Petter \AA{}sen}$^{1}$, 
    	Andreas Austeng$^{2}$ and
    	Sverre Holm$^{1,2}$
} 
{
	$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
    	$^{2}$Department of Informatics, University of Oslo, Oslo, Norway
}
\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}
%\input{./CaponMovingObjects/CaponMovingObjects}
%\newpage\addcontentsline{toc}{section}{References}
\putbib[bibCaponMovingObjects]
\end{bibunit}

%%%% Paper 4 %%%%
\begin{bibunit}[ieeetr]
\CustChapterPaper{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}
\runningtitle{Adaptive Volume Rendering of Cardiac Ultrasound}
\authors{
	\textbf{Jon Petter \AA{}sen}$^{1}$,
	Erik Steen$^2$,
	Gabriel Kiss$^{3,4}$,
	Anders Thorstensen$^{1,5}$,
	Stein Inge Rabben$^2$
}
{
$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
$^{2}$ GE Vingmed Ultrasound, Horten, Norway\\
$^{3}$ Norwegian University of Science and Technology, ISB - Department of Circulation and Medical Imaging, Trondheim, Norway\\
$^{4}$ St. Olavs Hospital, Trondheim, Norway\\
$^{5}$ Department of Cardiology, St. Olavs Hospital, Trondheim, Norway
}
\noindent \textit{Proc. SPIE Medical Imaging 2012 \todo{add pages}}
%\input{./AdaptiveVolumeRendering/AdaptiveVolumeRendering}
%\newpage\addcontentsline{toc}{section}{References}
\putbib[bibAdaptiveVolumeRendering]
\end{bibunit}

%%%% Paper 5 %%%%
\begin{bibunit}[ieeetr]
\CustChapterPaper{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}
\runningtitle{Huygens on speed}
\authors{\textbf{Jon Petter \AA{}sen}$^1$ and Sverre Holm$^{1,2}$}
{
	$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
    	$^{2}$Department of Informatics, University of Oslo, Oslo, Norway
}
\noindent \textit{Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}}
%\input{./Hos/Hos}
%\newpage\addcontentsline{toc}{section}{References}
\putbib[bibHos]
\end{bibunit}

\appendix

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TITLE PAGE
\author{Jon Petter \AA{}sen}
\title{%Integrated Ultrasound Visualization and Formation
%Real-Time Adaptive Beamforming, Visualization, and Simulation 
%Real-Time Adaptive Ultrasound Algorithms
%Accelerating adaptive ultrasound algorithms using graphics processing units 
%Accelerating adaptive ultrasound algorithms with graphics processing units 
Accelerating adaptive ultrasound algorithms by means of general-purpose programming on graphics processing units 
%Accelerating adaptive ultrasound algorithms by general purpose programming on graphics processing units 
%Accelerating ultrasound algorithms with graphics processing units 
%Accelerating ultrasound algorithms using graphics processing units 
\vspace{0.5cm}}
\department{Department of Circulation and Medical Imaging}

@@

% symbol list
\usepackage{nomencl}
\makenomenclature
\renewcommand{\nomname}{Symbols and Abbreviations}
\newcommand*{\nom}[2]{#1\nomenclature{#1}{#2}}

\begin{document}

\date{\today}
\pagestyle{empty} \maketitle

@@

This thesis is based on the following five papers, referred to in the text by their Roman numerals (I-V): 
\begin{enumerate}[I]
\renewcommand\labelenumi{\bfseries\theenumi}
	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "An optimised GPU implementation of the MVDR beamformer for active sonar imaging", {\it  IEEE Transactions on Oceanic Engineering, submitted}.
	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon beamforming on a GPU for real-time cardiac ultrasound imaging", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, no. 1, pp. 76-85, Jan 2014}.
 	\item \textbf{J.\:P.\:\AA{}sen}, A.\:Austeng and S.\:Holm, "Capon beamforming and moving objects - An analysis of lateral shift-invariance", {\it IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}.
	\item \textbf{J.\:P.\:\AA{}sen}, E.\:Steen, G.\:Kiss, A.\:Thorstensen and S.\:I.\:Rabben, "Adaptive volume rendering of cardiac 3D ultrasound images - utilizing blood pool statistics", {\it Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008.}.
	\item \textbf{J.\:P.\:\AA{}sen} and S.\:Holm, "Huygens on speed: Interactive simulation of ultrasound pressure fields", {\it Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646}.
\end{enumerate}

@@

	\item \textbf{J.\:P.\:\AA{}sen}, J.\:I.\:Buskenes, C.-I.\:C.\:Nilsen, A.\:Austeng and S.\:Holm, "Implementing Capon Beamforming on the GPU for Real-Time Cardiac Ultrasound Imaging", {\it Proc. IEEE Ultrasonics Symposium 2012}, pp. 2133-2136.

	\item J.\:I.\:Buskenes, \textbf{J.\:P.\:\AA{}sen}, C.-I.\:C.\:Nilsen and A.\:Austeng, "Adapting the Minimum variance beamformer to a graphics processing unit for active sonar imaging systems", {\it Proc. Meetings on Acoustics 2013} vol. 13.

	\item G.\:Kiss, E.\:Steen, \textbf{J.\:P.\:\AA{}sen} and H.\:Torp, "GPU volume rendering in 3D echocardiography: real-time pre-processing and ray-casting", {\it Proc. IEEE Ultrasonics Symposium 2010}, pp. 193-196.

@@

	\item E.\:S.\:Br\o{}nstad, \textbf{J.\:P.\:\AA{}sen}, H.\:Torp and G.\:Kiss, "Visibility driven visualization of 3D cardiac ultrasound data on the GPU", {\it Proc. IEEE Ultrasonics Symposium 2012}, pp. 2651-2654.
\end{enumerate}

@@

\cleardoublepage
\addcontentsline{toc}{chapter}{Symbols and Abbreviations}
\printnomenclature

\cleardoublepage
\setcounter{page}{1}
\renewcommand{\thepage}{\arabic{page}}

%\renewcommand*\thesection{\arabic{section}}

%%%% CHAPTER 2 %%%%
\begin{bibunit}[ieeetr]
\CustChapter{Background}
\runningtitle{Background}
\input{./Background/background}
\newpage\addcontentsline{toc}{section}{References}
\putbib[./Background/background]
\end{bibunit}

%\setcounter{section}{0}
%\setcounter{chapter}{1}

%%%% CHAPTER 1 %%%%
\CustChapter{Introduction} 
\runningtitle{Introduction}
\input{./Introduction/introduction}

@@

%%%% Paper 1 %%%%
\begin{bibunit}[ieeetr]
\CustChapterPaper{An Optimised GPU Implementation of the MVDR Beamformer for Active Sonar Imaging}
\runningtitle{GPU MVDR Beamformer for Active Sonar Imaging}
\authors{
	Jo Inge Buskenes$^{1}$, 
    	\textbf{Jon Petter \AA{}sen}$^{2}$, 
    	Carl-Inge C. Nilsen$^{1}$ and
    	Andreas Austeng$^{1}$
} 
{
	$^{1}$Department of Informatics, University of Oslo, Oslo, Norway\\
    	$^{2}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway
}
\noindent \textit{IEEE Transactions on Oceanic Engineering, submitted}.
%\newpage\input{./CaponSonarGPU/CaponSonarGPU}
%\newpage\addcontentsline{toc}{section}{References}
%\putbib[./CaponSonarGPU/bibCaponSonarGPU]
\end{bibunit}

%%%% Paper 2 %%%%
\begin{bibunit}[ieeetr]
\CustChapterPaper{Implementing Capon Beamforming on a GPU for Real-Time Cardiac Ultrasound Imaging}
\runningtitle{GPU Capon Beamformer for Cardiac Ultrasound Imaging}
\authors{
	\textbf{Jon Petter \AA{}sen}$^{1}$,
    	Jo Inge Buskenes$^{2}$,  
    	Carl-Inge C. Nilsen$^{2}$,
    	Andreas Austeng$^{2}$ and
    	Sverre Holm$^{1,2}$
} 
{
	$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
    	$^{2}$Department of Informatics, University of Oslo, Oslo, Norway
}
\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, vol. 61, pp. 76-85, Jan 2014}.
%\input{./CaponCardiacGPU/CaponCardiacGPU}
%\newpage\addcontentsline{toc}{section}{References}
%\putbib[./CaponCardiacGPU/bibCaponCardiacGPU]
\end{bibunit}

%%%% Paper 3 %%%%
\begin{bibunit}[ieeetr]
\CustChapterPaper{Capon Beamforming and Moving Objects - An Analysis of Lateral Shift-Invariance}
\runningtitle{Capon Beamforming and Moving Objects}
\authors{
	\textbf{Jon Petter \AA{}sen}$^{1}$, 
    	Andreas Austeng$^{2}$ and
    	Sverre Holm$^{1,2}$
} 
{
	$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
    	$^{2}$Department of Informatics, University of Oslo, Oslo, Norway
}
\noindent \textit{IEEE Transactions on Ultrasonics, Ferroelectrics, and Frequency Control, submitted}.
%\input{./CaponMovingObjects/CaponMovingObjects}
%\newpage\addcontentsline{toc}{section}{References}
%\putbib[./CaponMovingObjects/bibCaponMovingObjects]
\end{bibunit}

%%%% Paper 4 %%%%
\begin{bibunit}[ieeetr]
\CustChapterPaper{Adaptive Volume Rendering of Cardiac 3D Ultrasound Images - Utilizing Blood Pool Statistics}
\runningtitle{Adaptive Volume Rendering of Cardiac Ultrasound}
\authors{
	\textbf{Jon Petter \AA{}sen}$^{1}$,
	Erik Steen$^2$,
	Gabriel Kiss$^{3,4}$,
	Anders Thorstensen$^{1,5}$,
	Stein Inge Rabben$^2$
}
{
$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
$^{2}$ GE Vingmed Ultrasound, Horten, Norway\\
$^{3}$ Norwegian University of Science and Technology, ISB - Department of Circulation and Medical Imaging, Trondheim, Norway\\
$^{4}$ St. Olavs Hospital, Trondheim, Norway\\
$^{5}$ Department of Cardiology, St. Olavs Hospital, Trondheim, Norway
}
\noindent \textit{Proc. SPIE Medical Imaging 2012, vol. 8320, pp. 832008.}
%\input{./AdaptiveVolumeRendering/AdaptiveVolumeRendering}
%\newpage\addcontentsline{toc}{section}{References}
%\putbib[./AdaptiveVolumeRendering/bibAdaptiveVolumeRendering]
\end{bibunit}

%%%% Paper 5 %%%%
\begin{bibunit}[ieeetr]
\CustChapterPaper{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}
\runningtitle{Huygens on speed}
\authors{\textbf{Jon Petter \AA{}sen}$^1$ and Sverre Holm$^{1,2}$}
{
	$^{1}$Medical Imaging Lab (MI-Lab), Norwegian University of Science and Technology, Trondheim, Norway\\
    	$^{2}$Department of Informatics, University of Oslo, Oslo, Norway
}
\noindent \textit{Proc. IEEE Ultrasonics Symposium 2012, pp. 1643-1646}.
%\input{./Hos/Hos}
%\newpage\addcontentsline{toc}{section}{References}
%\putbib[./Hos/bibHos]
\end{bibunit}

%\appendix
